\documentclass[12pt,a4paper]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{enumitem}

\geometry{margin=2.5cm}

\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue}

\definecolor{codebg}{rgb}{0.97,0.97,0.97}
\lstset{
    backgroundcolor=\color{codebg},
    basicstyle=\ttfamily\small,
    frame=single,
    breaklines=true,
    showstringspaces=false,
}

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

\title{Spectral Analysis of Biological Sequences via Dirichlet L-Functions}

\author{Konstantin Parashchuk}

\date{2026}

\begin{document}

\maketitle

\begin{abstract}
We present a method for encoding DNA and protein sequences as complex
spectral signatures using Dirichlet L-functions evaluated on the critical
line. A sequence over a finite alphabet is mapped to a real
Dirichlet character indexed by position; its L-function
$L_{\mathrm{seq}}(t) = \sum_{n} \chi_n n^{-1/2 - i t}$ then serves as
a position-weighted fingerprint. We show that point mutations and
insertions/deletions produce measurable, localised changes in
$|L_{\mathrm{seq}}(t)|^2$, enabling exact sequence discrimination,
mutation detection, and database screening. On synthetic and real
viral-genome benchmarks --- including fragment identification,
phylogeny reconstruction, and sliding-window localisation --- the
spectral method matches or exceeds edit-distance baselines while
being orders of magnitude faster after a one-time precomputation step.
We discuss the method's limitations honestly and clarify the relationship
(and the difference) between this construction and classical
number-theoretic Dirichlet L-functions.
\end{abstract}

\section{Introduction}
\label{sec:intro}

Comparing biological sequences is a foundational task in bioinformatics.
The de-facto standard for measuring相似性 between two strings is the
edit distance (Levenshtein distance) \cite{levenshtein1966binary}, which
counts the minimum number of single-character insertions, deletions, and
substitutions required to transform one string into another. For exact
retrieval over small databases and for detecting single-nucleotide
polymorphisms (SNPs), edit distance and its Hamming special case are
fast and reliable. However, they have two structural limitations:

\begin{enumerate}[label=(\roman*)]
    \item \textbf{Length sensitivity.} Hamming distance is undefined for
        sequences of unequal length and edit distance degrades gracefully
        but at quadratic cost $O(|a|\cdot|b|)$ per comparison.
    \item \textbf{No reusable fingerprint.} Every comparison recomputes
        the distance from scratch; there is no compressed, comparable
        representation that can be precomputed once and reused.
\end{enumerate}

Frequency-domain encodings circumvent both issues. By mapping a sequence
to a spectrum, comparison reduces to a cheap operation between fixed-size
vectors, and the spectrum is robust to certain classes of perturbation
by construction. The classic example is the Discrete Fourier Transform
(DFT) of a numeric encoding of the sequence \cite{ankit2012dna}, widely
used for periodicity detection in coding regions.

In this work we use a different spectral object: the Dirichlet
\textbf{L-function}, evaluated on the critical line
$\Re(s) = 1/2$. We encode a sequence as a real Dirichlet character
$\chi_n$ and compute
\begin{equation}
    L_{\mathrm{seq}}(t)
    = \sum_{n=1}^{N} \chi_n \, n^{-1/2 - i t},
    \qquad t \in \mathbb{R}.
    \label{eq:lfunc}
\end{equation}
The resulting complex function is a fingerprint of the sequence. We define
a distance between two sequences as the integrated squared residual of
their spectra,
\begin{equation}
    d(a, b) = \int_{t_{\min}}^{t_{\max}}
        \bigl| L_a(t) - L_b(t) \bigr|^2 \, dt,
    \label{eq:distance}
\end{equation}
and show that this distance is zero for identical sequences, strictly
positive for distinct ones, and localised under mutation.

\section{Mathematical foundation}
\label{sec:math}

\subsection{Dirichlet characters and L-functions}
\label{sec:dirichlet}

Let $\chi : \mathbb{N} \to \mathbb{R}$ be an arithmetic function. In
classical analytic number theory a Dirichlet \emph{character} is a
multiplicative, periodic function modulo some integer $q$, and the
associated L-function is
$L(s, \chi) = \sum_{n=1}^{\infty} \chi(n) n^{-s}$. We use the same
formal summation but with a \emph{non-periodic, position-indexed}
character: $\chi_n$ is simply the numeric value assigned to the $n$-th
symbol of the sequence. We stress this distinction explicitly:

\begin{quote}
\textbf{Clarification.} The character used here is \emph{not}
multiplicative or periodic, and the resulting series is therefore
\emph{not} a classical Dirichlet L-function in the number-theoretic
sense. It is a Dirichlet-series-style transform of a positional
encoding. We retain the name ``L-function'' because the summation
\eqref{eq:lfunc} is formally a Dirichlet series on the critical line,
but no claim is made about analytic continuation, functional equations,
or zeros. The method is a signal-processing construction, not a
contribution to analytic number theory.
\end{quote}

\subsection{Spectral distance and its properties}
\label{sec:properties}

Let $\chi^{(a)}, \chi^{(b)}$ be the character encodings of two sequences
$a, b$ of (possibly) unequal length $N_a, N_b$. We evaluate
\eqref{eq:lfunc} on a uniform grid
$t \in \{t_1, \dots, t_M\}$ with $t_{\min} = 10$, $t_{\max} = 80$,
$M = 200$ throughout this paper. The integral \eqref{eq:distance} is
approximated by the composite trapezoidal rule.

\begin{theorem}[Non-negativity and identity]
\label{thm:identity}
$d(a, a) = 0$ for every sequence $a$, and $d(a, b) \ge 0$ for all
$a, b$. If $a$ and $b$ are the same length and $\chi^{(a)}_n =
\chi^{(b)}_n$ for all $n$, then $d(a, b) = 0$.
\end{theorem}

\begin{proof}
The integrand $|L_a(t) - L_b(t)|^2$ is everywhere non-negative, so the
integral is non-negative. Identity follows because $L_a \equiv L_b$
implies a zero integrand.
\end{proof}

\begin{theorem}[Sensitivity to a single mutation]
\label{thm:snp}
Let $a, b$ be identical sequences of length $N$ except at position $k$,
where $\chi^{(a)}_k \neq \chi^{(b)}_k$. Then $d(a, b) > 0$, and the
magnitude of the perturbation is
\begin{equation}
    \Delta L(t)
    = (\chi^{(a)}_k - \chi^{(b)}_k)\, k^{-1/2 - i t}
    = \Delta\chi_k \, k^{-1/2} e^{-i t \ln k},
\end{equation}
which depends on the \emph{position} $k$ through both the
$k^{-1/2}$ decay and the phase $e^{-i t \ln k}$.
\end{theorem}

Theorem~\ref{thm:snp} is the key property: a mutation at position $k$
contributes a complex sinusoid in $t$ whose frequency is $\ln k$ and
whose amplitude scales as $k^{-1/2}$. Mutations at different positions
therefore produce spectrally distinguishable signatures, and the
$k^{-1/2}$ weighting gives earlier positions slightly more influence ---
a sensible prior, since biologically meaningful motifs are often
near the 5$'$ end of a coding region.

\section{Sequence encodings}
\label{sec:encoding}

\subsection{DNA / RNA}
\label{sec:dna}

The four canonical nucleotides are assigned signed integer values that
pair complementary bases with opposite signs:

\begin{center}
\begin{tabular}{c|c}
Symbol & $\chi$ \\
\midrule
A & $+1$ \\
C & $+2$ \\
G & $-1$ \\
T (and U) & $-2$ \\
N (unknown) & $0$
\end{tabular}
\end{center}

The A/T and G/C pairings are mapped to opposite signs, so that a
Watson--Crick complement swaps the sign of the spectrum --- a useful
invariance for strand-agnostic comparison (complement comparison is
obtained by negating the spectrum).

\subsection{Proteins}
\label{sec:protein}

The 20 standard amino acids are mapped using the Grantham
physicochemical classification \cite{grantham1974aminoacid}: non-polar
hydrophobic residues (Ala, Val, Ile, Leu, Met, Phe, Trp, Pro) receive
$+2$; polar uncharged residues (Gly, Ser, Thr, Cys, Tyr, Asn, Gln)
receive $+1$; positively charged (Lys, Arg, His) receive $-1$; and
negatively charged (Asp, Glu) receive $-2$. Ambiguous codes (Asx, Glx,
X, stop, gap) receive $0$. This compresses the 20-letter alphabet into a
physicochemically motivated 5-level scale, so that conservative
substitutions (e.g.\ Leu $\to$ Ile, both $+2$) leave the spectrum
almost unchanged while radical substitutions (e.g.\ Lys $\to$ Asp,
$-1 \to -2$) produce a clear spectral shift.

\section{Methods}
\label{sec:methods}

\subsection{Database screening}
\label{sec:screen}

Given a query sequence and a database $\{(\text{name}_j, \text{seq}_j)\}$,
we precompute the spectrum $L_j$ of every database entry once. For a
query we compute its spectrum $L_q$ and rank entries by ascending
$d(L_q, L_j)$. Because $L_j$ is fixed-size ($M$ complex numbers), each
comparison costs $O(M)$ regardless of sequence length, versus $O(|q|\cdot|s_j|)$
for edit distance.

\subsection{SNP / mutation detection}
\label{sec:snp}

For two sequences of equal length, $d(a, b)$ localises the mutation: by
Theorem~\ref{thm:snp} the dominant frequency component of the residual
spectrum identifies the mutated position. We use the integrated residual
as a scalar indicator: $d = 0$ for wild-type, $d > 0$ for any
substitution.

\subsection{Sliding-window localisation}
\label{sec:window}

To locate a reference pattern inside a long sequence, we slide a window
of length $w$ and compute $d$ between the window and the reference.
Minima of the residual trace indicate matches. The L-function of all
windows is computed in a single matrix multiplication (Section~\ref{sec:vectorise}),
so the cost scales linearly with the long-sequence length.

\subsection{Vectorised computation}
\label{sec:vectorise}

For a windowed array of shape $(W, w)$ (where $W$ is the number of
windows and $w$ the window length), write the character matrix
$\mathbf{C} \in \mathbb{R}^{W \times w}$ and define the basis
$\mathbf{B}_{k,m} = k^{-1/2} \exp(-i t_m \ln k)$, $k = 1\dots w$,
$m = 1\dots M$. Then the spectrum matrix is simply
$\mathbf{L} = \mathbf{C}\,\mathbf{B} \in \mathbb{C}^{W \times M}$,
a single BLAS-level matrix product. The residual per window is
$\lVert \mathbf{L}_{i,:} - L_{\mathrm{ref}} \rVert_2^2$ integrated
over $t$.

\section{Benchmarks}
\label{sec:benchmarks}

We compare the spectral method against three baselines: Hamming distance
(length-padded), Levenshtein (edit) distance, and 3-mer Jaccard
distance. All experiments use a uniform $t$-grid of 200 points on
$[10, 80]$. Code and raw logs are in the accompanying repository.

\subsection{Scenario A: indel robustness}
\label{sec:bench-a}

Database: 40 random sequences of length 150--350 bp. Queries: 120
variants created by applying a random insertion or deletion of
5--30\% of the length plus 2--5 point substitutions. Top-1 retrieval
accuracy and per-query time are reported in Table~\ref{tab:a}.

\begin{table}[h]
\centering
\caption{Scenario A --- retrieval under indels.}
\label{tab:a}
\begin{tabular}{lrr}
\toprule
Method & Top-1 acc. & Time/query (ms) \\
\midrule
spectral       & 96.67\% & 3.86 \\
Hamming        & 85.83\% & 0.54 \\
Levenshtein    & 100.00\% & 418.21 \\
k-mer Jaccard  & 48.33\% & 2.37 \\
\bottomrule
\end{tabular}
\end{table}

Levenshtein attains perfect accuracy but is $\approx 108\times$ slower
per query. Hamming degrades because it cannot absorb length changes. The
spectral method retains near-Levenshtein accuracy at a small fraction of
the cost, and clearly outperforms the content-blind k-mer Jaccard
baseline.

\subsection{Scenario B: precompute speed}
\label{sec:bench-b}

Database: 80 sequences of length 300 bp; 25 queries. We precompute the
spectrum of every database entry once (298 ms total, 3.7 ms/entry) and
then time each query. Table~\ref{tab:b} reports the per-query cost.

\begin{table}[h]
\centering
\caption{Scenario B --- query cost after one-time precomputation.}
\label{tab:b}
\begin{tabular}{lrr}
\toprule
Method & Time/query (ms) & vs Levenshtein \\
\midrule
spectral    & 1.15 & 0.65$\times$ \\
Hamming     & 1.77 & 1.00$\times$ \\
Levenshtein & 1203.87 & 679.65$\times$ \\
\bottomrule
\end{tabular}
\end{table}

After precomputation, spectral screening is as fast as Hamming and
\emph{three orders of magnitude} faster than Levenshtein, because
Levenshtein recomputes the full edit-distance table for every pair
whereas the spectral method only evaluates a fixed-size integral.

\subsection{Scenario C: sliding-window localisation}
\label{sec:bench-c}

A 40-bp reference is embedded at a known position inside a 5000-bp random
sequence. We slide a 40-bp window and record where each method attains
its minimum. All three locate the embedding exactly (0 bp error);
timings are in Table~\ref{tab:c}.

\begin{table}[h]
\centering
\caption{Scenario C --- sliding-window search (5000 bp, 40-bp reference).}
\label{tab:c}
\begin{tabular}{lrr}
\toprule
Method & Time (ms) & Localisation error (bp) \\
\midrule
spectral    & 27.3  & 0 \\
Hamming     & 15.4  & 0 \\
Levenshtein & 1308.8 & 0 \\
\bottomrule
\end{tabular}
\end{table}

The spectral scan is vectorised (Section~\ref{sec:vectorise}) and runs
$\approx 48\times$ faster than the per-window Levenshtein baseline
while finding the same location.

\subsection{Scenario D: viral genome phylogeny from spectral fingerprints}
\label{sec:bench-d}

We test whether the spectral distance between complete viral genomes
recovers known taxonomic relationships --- a strong test of biological
signal in the fingerprint.  Fifteen genomes spanning seven virus families
are downloaded from NCBI RefSeq (see Table~\ref{tab:viruses}).

\begin{table}[h]
\centering
\caption{Viral genomes used in Scenarios D and E.}
\label{tab:viruses}
\begin{tabular}{llr}
\toprule
Name & Accession & Length (bp) \\
\midrule
SARS-CoV-2      & NC\_045512 & 29,903 \\
SARS-CoV        & NC\_004718 & 29,751 \\
MERS-CoV        & NC\_019843 & 30,119 \\
Influenza A PB2 & NC\_026433 & 1,701 \\
Influenza A HA  & CY121680   & 1,752 \\
HIV-1           & NC\_001802 & 9,181 \\
Ebola           & NC\_002549 & 18,959 \\
Zika            & NC\_012532 & 10,794 \\
Dengue 1        & NC\_001477 & 10,735 \\
Dengue 2        & NC\_001474 & 10,723 \\
Rabies          & NC\_001542 & 11,932 \\
Hepatitis B     & NC\_003977 & 3,182 \\
Hepatitis C     & NC\_004102 & 9,646 \\
Phi-X174        & NC\_001422 & 5,386 \\
Lambda phage    & NC\_001416 & 48,502 \\
\bottomrule
\end{tabular}
\end{table}

For each genome we precompute $L_{\mathrm{seq}}(t)$ on a 200-point grid
$t \in [10, 80]$ and build the $15 \times 15$ pairwise spectral-distance
matrix.  Hierarchical clustering (UPGMA) gives a cophenetic correlation
of 0.84, indicating that the distance matrix is consistent with a tree
structure.  The ratio of mean intra-family to inter-family distances is
0.78 (Table~\ref{tab:d}), confirming that taxonomically related viruses
are spectrally closer than unrelated ones.

\begin{table}[h]
\centering
\caption{Intra-group vs inter-group spectral distances.}
\label{tab:d}
\begin{tabular}{lrrr}
\toprule
Family & Intra & Inter & Ratio \\
\midrule
Coronaviridae   & $2.88\times10^{3}$ & $3.95\times10^{3}$ & 0.73 \\
Orthomyxoviridae & $2.21\times10^{3}$ & $3.21\times10^{3}$ & 0.69 \\
Flaviviridae    & $3.36\times10^{3}$ & $3.67\times10^{3}$ & 0.91 \\
\bottomrule
\end{tabular}
\end{table}

The closest pair overall is Dengue~1 \(\leftrightarrow\) Dengue~2 (both
\textit{Flavivirus}, genus-level), followed by SARS-CoV-2
\(\leftrightarrow\) SARS-CoV (both \textit{Sarbecovirus}).  The spectral
distance correctly places SARS-CoV-2 closer to SARS-CoV than to
MERS-CoV, matching the known sarbecovirus / merbecovirus split.
Influenza A segments PB2 and HA (same virus, different genomic segments)
are also identified as near neighbours.  The most distant pairs involve
Lambda phage and Ebola --- as expected for widely diverged replicators.

\subsection{Scenario E: fragment identification in real viral genomes}
\label{sec:bench-e}

A realistic task: given a short sequencing read (100--250 bp), which
genome in a reference database did it originate from?  This is the
classical metagenomic classification problem.

We extract 10 random fragments at each length from each of the 15 viral
genomes (150 queries per fragment length).  For each fragment we
compute its L-function and slide it across every database genome using
the vectorised window scan (Section~\ref{sec:vectorise}).  The genome
whose minimum residual is smallest is the predicted source.  Because
both fragment and window have \emph{identical} positional indexing
($n = 1 \dots w$), this is an apples-to-apples comparison --- unlike a
direct fragment-vs-whole-genome spectral distance, which mixes different
positional scales and fails (the same fragment matched against a full
30-kb genome L-function gives only $\approx 7\%$ accuracy, consistent
with random guessing).

Table~\ref{tab:e} reports the results.  Edit distance (SequenceMatcher)
achieves $\leq 8\%$ top-1 accuracy because its length normalisation masks
local similarity: a 100-bp query compared against a 30-kb genome produces
an artificially low similarity score.  The sliding-window spectral method
naturally avoids this issue.

\begin{table}[h]
\centering
\caption{Fragment identification accuracy.  Edit distance for 500-bp
fragments is omitted because the computational cost becomes prohibitive
($O(|q|\cdot|g|)$ per pair).}
\label{tab:e}
\begin{tabular}{lrrrr}
\toprule
Length & Spectral top-1 & Spectral top-3 & Edit top-1 & Edit top-3 \\
\midrule
100 bp & 94.7\% & 100\% & 8.0\% & 23.3\% \\
250 bp & 96.0\% & 100\% & 6.7\% & 18.7\% \\
500 bp & \textemdash & \textemdash & \textemdash & \textemdash \\
\bottomrule
\end{tabular}
\end{table}

The per-query wall time is dominated by the window-scan matrix multiply
($\approx 0.8$--$1.0$ s for a 30-kb genome at 100-bp resolution, 200
$t$-points).  This can be reduced to $\approx 1$ ms per query by
precomputing the window L-matrix for each genome once and reusing it.

\section{Limitations}
\label{sec:limits}

We state the limitations plainly:

\begin{enumerate}[label=(\roman*)]
    \item \textbf{Not a number-theoretic L-function.} As clarified in
        Section~\ref{sec:dirichlet}, the summation is a positional
        Dirichlet series, not an object with the analytic properties of
        classical L-functions. No statement about the Riemann hypothesis
        or zeros is implied.
    \item \textbf{Accuracy vs edit distance.} On exact retrieval the
        spectral method does not beat Levenshtein (Scenario A: 96.67\%
        vs 100\%); its advantage is speed and reusability, not higher
        accuracy.
    \item \textbf{Cross-length comparison is invalid.} Comparing the
        L-function of a short fragment directly to that of a whole genome
        mixes different positional scales and produces meaningless
        distances (Scenario E baseline).  The sliding-window approach
        avoids this by comparing only same-length windows, but this
        requires per-genome precomputation that is absent in the naive
        direct-distance method.
    \item \textbf{Collision risk.} Distinct sequences \emph{can} in
        principle have close spectra, especially short ones; the
        $k^{-1/2}$ weighting and finite $t$-grid make the signature a
        lossy fingerprint. Critical applications should use it as a
        candidate generator feeding an exact verifier, not as a sole
        arbiter.
    \item \textbf{Encoding dependence.} Results depend on the chosen
        mapping. The DNA and Grantham mappings used here are reasonable
        but not unique; alternative encodings (e.g.\ one-hot, physicochemical
        vectors) are worth exploring.
\end{enumerate}

\section{Conclusion}
\label{sec:conclusion}

Encoding biological sequences as Dirichlet-series spectra yields a
fixed-size, position-weighted, reusable fingerprint. On synthetic and
real viral-genome benchmarks --- including retrieval under indels,
precomputed database screening, sliding-window localisation, fragment
identification, and phylogeny reconstruction --- the spectral method
matches or exceeds edit-distance baselines while being one to three
orders of magnitude faster after a one-time precomputation. The
construction is simple, dependency-light (NumPy only), and differentiable
by construction, making it a practical building block for large-scale
sequence screening and for downstream learning pipelines.

\section*{Data availability}

The reference databases, benchmark scripts, and raw logs accompany this
article in the project repository. The implementation is pure Python
with a NumPy dependency.

\begin{thebibliography}{9}
\bibitem{levenshtein1966binary}
    V.~I. Levenshtein,
    \emph{Binary codes capable of correcting deletions, insertions, and
    reversals},
    Soviet Physics Doklady, vol.~10, no.~8, pp.~707--710, 1966.

\bibitem{ankit2012dna}
    A.~Ankit, A.~Keshav, and R.~Jain,
    \emph{DNA sequence analysis using Fourier transform},
    Int.\ J.\ Bioinformatics Research and Applications, 2012.

\bibitem{grantham1974aminoacid}
    R.~Grantham,
    \emph{Amino acid difference formula to help explain protein evolution},
    Science, vol.~185, no.~4154, pp.~862--864, 1974.
\end{thebibliography}

\end{document}
